### Lecture 9 code: Environments
### Debashis Ghosh

op <- par()
options(width=80)
emptyenv()
rm(list=ls())

#install.packages("ggplot2")
#install.packages("rlang") # this actually just needed to be updated for ggplot2
#install.packages("pryr")
library(ggplot2)
library(pryr)

# generate a data frame called df

df <- data.frame(gp = factor(rep(letters[1:3], each = 10)),
                 y = rnorm(30))
head(df)

# Illustrate environments

?ggplot

# here is a useful tutorial on how ggplot2 works
# https://mgimond.github.io/ES218/Week04c.html
                 
# Illustrate scoping rules    
# Generate some sample data, then compute mean and standard deviation
# in each group. This comes from 
# https://ggplot2.tidyverse.org/reference/ggplot.html

```{r}
df <- data.frame(
  gp = factor(rep(letters[1:3], each = 10)),
  y = rnorm(30)
)
ds <- plyr::ddply(df, "gp", plyr::summarise, mean = mean(y), sd = sd(y))

# The summary data frame ds is used to plot larger red points on top
# of the raw data. Note that we don't need to supply `data` or `mapping`
# in each layer because the defaults from ggplot() are used.
ggplot(df, aes(gp, y)) +
  geom_point() +
  geom_point(data = ds, aes(y = mean), 
             colour = 'red', size = 3)
search()
ddply
```
              
# Alternatively we can fully specify the plot in each layer. This
# is not useful here, but can be more clear when working with complex
# mult-dataset graphics
```{r}
ggplot() +
  geom_point(data = df, aes(gp, y)) +
  geom_point(data = ds, aes(gp, mean), colour = 'red', size = 3) +
  geom_errorbar(
    data = ds,
    aes(gp, mean, ymin = mean - sd, ymax = mean + sd),
    colour = 'red',
    width = 0.4
  )
```
# Do a search to identify the tree of the environment    
```{r}
search()

# Illustrate the functions

ls.str("package:methods")

parent.env(".GlobalEnv") ### now it is a string

environment(ggplot2::ggplot)

parent.env(as.environment("package:stats"))

identical(parent.env(as.environment("package:stats")),globalenv())
```

# Illustrate where

*** cannot use the == or = ***
```{r}
where <- function(name, env = parent.frame()) {
  if (identical(env, emptyenv())) {
    # Base case 
    stop("Can't find ", name, call. = FALSE)
    
  } else if (exists(name, envir = env, inherits = FALSE)) {
    # Success case
    env
    
  } else {
    # Recursive case
    where(name, parent.env(env)) ### where function is calling the where function, call it self.
    
  }
}
```


#  
environment(mean)   # gives the binding environment
where("mean") # gives the enclosing environment
environment(ggplot)
where("ggplot")
where("aes")

# Functions #
y <- 1
f <- function(x) x + y
environment(f)
environment(y)
ls()


# calling environments #
h <- function () {
  x <- 10
  function () {
    x
  }
}
i <- h()
x <- 20
# what will i() return?
i()

# x is 10 in the environment where h() is defined and 20
# in the environment where h() is called
# we can access this using the parent.frame()
f2 <- function () {
  x <- 10
  function () {
    def <- get("x", environment())
    cll <- get("x", parent.frame())
    list(defined = def, called = cll)
  }
}
g2 <- f2()
x <- 20
str(g2())


