---
title: "Objects in R"
author: "Debashis Ghosh"
date: "9/24/2019"
output: html_document
---

# Objects
  + Another key feature of any programming language
  + One prototype: C++ (object-oriented (OO) programming)
  + Key concepts: class and method
  + Class defines the behavior of objects by 
describing their attributes and relationship to other classes
  + Methods then operate on the particular class
  + Classes organized in a hierarchy so that if a particular method does not exist, the ``parent" of the object are found and then the method of the parent class is used (this is called **inheritance**)


# Objects in R 
  + S3, S4, and Reference classes (RC)
  + S3 implements generic-function OO and is diferent from other 
languages (e.g., C++).  Has no definition of classes
  + S4 has more formal class definitions 
  + Reference classes (RC) implements message-passing 
object-oriented programming, so methods belong to classes
not functions


# Base types
  + An underlying C-level types that underlie the OO systems. 
  + Idea: for every R object, there is a C structure (struct)
that describes how the object is stored in memory.   
  + C structure includes contents of the object, information
needed for memory management, and a **type**
  + Every object is built on top of the base types.
```{r}
x<-sample(c("A","C","T","G"), size=1000, replace=T)
mean(x)
today <- Sys.Date()
mean(today)
```
  
# Encapsulation
  + Put all the functions of initialization and functions inside a 
  class
  + Example: https://www.kdnuggets.com/2019/08/object-oriented-programming-data-scientists-estimator.html
  
  python language
  self.a = a
  
# S3
  + Oldest object system in R
  + Most objects in the base and stat packages in R have this object type
  + ***No strict rules for S3 class definition***

```{r}
# create a list with required components
s <- list(name = "John", age = 21, GPA = 3.5)
# name the class appropriately
class(s) <- "student"
# That's it! we now have an object of class "student"
s
attr(s,"class")
attributes(s)
```

# Constructor
  + Formal subroutine used to construct an object
  
```{r}
# a constructor function for the "student" class
student <- function(n,a,g) {
# we can add our own integrity checks
if(g>4 || g<0)  stop("GPA must be between 0 and 4")
value <- list(name = n, age = a, GPA = g)
# class can be set using class() or attr() function
attr(value, "class") <- "student"
value
}

s <- student("Paul", 26, 3.7)
s <- student("John", 25, 3.6)
#s <- student("Paul", 26, 5)
s

```
```{r}
## generics 
mean.Date
```

# Functions using S3 objects

  + Methods belong to functions in S3, NOT to objects
  + Functions involving S3 will generally call the function **UseMethod()**
  + A function like **UseMethod()** figures out the proper function to call; this is called **method dispatch**
  + Some primitive functions don't use UseMethod();
they instead use internal functions in C to do method dispatch.
  + Given a class, the job of an S3 generic is to call the right S3 method.
  + S3 methods have the structure **generic.class()**

# Creating a new method and generic in S3


```{r}
methods(print)


# Now, create a print method to handle the student class

print.student <- function(obj) {
cat(obj$name, "\n") ## new line operator "\n" goes to next line; otherwise all the things will be in one line.
cat(obj$age, "years old\n") ## cat knits all the things in the () together.
cat("GPA:", obj$GPA, "\n")
}

print(s)
```


```{r}

# Create a generic function for getting a grade

grade <- function(obj) {
UseMethod("grade")         # Method dispatch
}


grade.default <- function(obj) {
cat("This is a generic function\n")
}

grade.student <- function(obj) {
cat("Your grade is", obj$GPA, "\n")
}

grade(s)
```

# S4 objects

  + Classes have formal definitions which describe their fields and inheritance structures (parent classes)
  + Method dispatch can be based on multiple arguments to a generic function, not just one.
  + There is a special operator, @, for extracting slots (aka fields) from an S4 object.
  + To identify S4 objects, one can use {\bf str()}, which will describes S4 as a "formal" class, {\bf isS4()} returns TRUE, and pryr::otype() returns "S4".

# Creating new methods and generics in S4
  + Key functions are ***setGeneric()*** (creates a new Generic) and setMethod(), which takes the name of the generic, 
the classes the method should be associated with, 
and a function that implements the method

```{r}
setClass("Person", representation(name = "character", age = "numeric"))
setClass("Employee", representation(boss = "Person"), contains = "Person")

paul <- new("Person",name="Paul",age=42)
paul
typeof(paul)
paul@age ### attributes @ the same as the $ in the list
paul@name

## wrong <- new("Person",name="Debashis",age="42")
## S4 is strongly typed; it is creating a new rule
john <- new("Employee", boss = paul, name="John", age=30)
john@boss
john
```


# Reference Classes (RC)

  + RC methods belong to objects, not functions
  *** Python methods ***
  + RC objects are mutable: the usual R copy-on-modify semantics do not apply
  + Reference classes are implemented using R code: they are a special S4 class that wraps around a certain environment
  + To create a new RC class, use **setRefClass()**
  + Use **new()** in addition to the object defined by setRefClass() to create the new objects
  + Define methods using **methods()** and inheritance using **contains()**
  
```{r,echo=T}

Account <- setRefClass("Account",
  fields = list(balance = "numeric"))
  
  a <- Account$new(balance = 100)
a$balance

b <- a
a$balance <- 0

b$balance

d <- a$copy()

d$balance

a$balance <- 150

d$balance

Account <- setRefClass("Account",
  fields = list(balance = "numeric"),
  methods = list(
    withdraw = function(x) {
      balance <- balance - x
    },
    deposit = function(x) {
      balance <<- balance + x ### globale assign operator
    }
  )
)

a <- Account$new(balance = 100); a$balance

a$deposit(100)
a$balance
a$withdraw(300)
a$balance
```

