---
title: 'Data Transformation lecture'
author: "Debashis Ghosh"
date: "October 22, 2019"
output:
  html_document: default
  fig_caption: yes
  number_sections: yes
  pdf_document: default
  word_document: default
---
#### This lecture is a followup from the previous one and is a deeper dive into the `tidyverse' packages in R.  Note: we are also making use of Chapter 5 of the wonderful website stat545.com.  

##### We will use the `dplyr` package from `tidyverse`. 
```{r, echo=T, message=F, warning=F}
#install.packages("tidyverse", dependencies = T)
library(tidyverse)
```

##### I will also use data from Gapminder at https://www.gapminder.org/data/.
```{r, echo=F, message=F, warning=F}
#install.packages("gapminder")
library(gapminder)
```

#### **Adding new variables with `mutate()`**
```{r, echo=T, message=T, warning=T}
gapm <- gapminder
newgap <- mutate(gapm,totgdp=pop*gdpPercap)
newgap
```

#### Try to lag change in life expectancy for each country
```{r, echo=T, message=T, warning=T}
tmp <- NULL
for (cnames in unique(as.character(gapminder$country))) {
  cd <- filter(gapminder,country == cnames )
  le <- select(cd,"lifeExp")   ### it is a list, so we have to unlist
  tmp <- c(tmp,lag(unlist(le),3))
}
gapminder$llifeExp <- tmp
plot(lifeExp ~ llifeExp,gapminder)
```

##### Let's say we want a quick look at aggregated statistics on a variable. One way to do this is by using: `cumsum()`, `cumprod()`, `cummin()`, `cummax()`; and `dplyr` provides `cummean()` for cumulative means.
```{r, echo=T, message=T, warning=T}
head(cummin(gapminder$lifeExp))
head(cummax(gapminder$lifeExp))
head(cummean(gapminder$lifeExp))
```

#### **Grouped summaries:** `summarize()`
##### The `summarize()` function is very powerful. It allows us to quickly summarize variables over meaningful groups without subsetting our data first (similar to `apply()` functions).
```{r, echo=T, message=T, warning=T}
by_country <- group_by(gapminder, country)
summarise(by_country, avgle = mean(lifeExp, na.rm = TRUE),avggdp 
          = mean(gdpPercap,na.rm=T))
```

##### Combining operations with the "pipe"

##### The code above requires creation of multiple new dataframes. If we use "pipe" notation, we can avoid this. The pipe notation refers to the %>% operator.  
```{r, echo=T, message=T, warning=T}
library(dplyr)
newc <- gapminder %>% 
  group_by(country) %>% 
  summarise(
    avgle = mean(lifeExp, na.rm = TRUE),
    avggdp = mean(gdpPercap,na.rm=T)
) 
newc
```

##### Quoting from the book: "This focuses on the transformations, not what’s being transformed, which makes the code easier to read. You can read it as a series of imperative statements: group, then summarise, then filter. As suggested by this reading, a good way to pronounce %>% when reading code is “then”. Behind the scenes, `x %>% f(y)` turns into `f(x, y)`, and `x %>% f(y) %>% g(z)` turns into `g(f(x, y), z)` and so on. You can use the pipe to rewrite multiple operations in a way that you can read left-to-right, top-to-bottom. We'’'ll use piping frequently from now on because it considerably improves the readability of code, and we’ll come back to it in more detail in pipes."


#### **Counts**
##### It is good to keep track of the group sizes when you're aggregating. You can do this using a count. I'll use the book's flights data for this example because the weather data are purposefully of equal group sizes.
```{r, echo=T, message=T, warning=T}
#install.packages('nycflights13')
library(nycflights13)

not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))

delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)
```
  
##### Some flights are delayed very long! But let's look at how many are delayed by delay time.
```{r, echo=T, message=T, warning=T}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

